Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2023-02-06 paper-1/manuscript.tex
Housekeeping
Rev abstract

%%%%%=============================================================================%%%%
%%%%  Remarks: This template is provided to aid authors with the preparation
%%%%  of original research articles intended for submission to journals published 
%%%%  by Springer Nature. The guidance has been prepared in partnership with 
%%%%  production teams to conform to Springer Nature technical requirements. 
%%%%  Editorial and presentation requirements differ among journal portfolios and 
%%%%  research disciplines. You may find sections in this template are irrelevant 
%%%%  to your work and are empowered to omit any such section if allowed by the 
%%%%  journal you intend to submit to. The submission guidelines and policies 
%%%%  of the journal take precedence. A detailed User Manual is available in the 
%%%%  template package for technical guidance.
%%%%%=============================================================================%%%%

@@

%% as per the requirement new theorem styles can be included as shown below
\theoremstyle{thmstyleone}%
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
\newtheorem{proposition}[theorem]{Proposition}% 
%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.

\theoremstyle{thmstyletwo}%
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%

\theoremstyle{thmstylethree}%
\newtheorem{definition}{Definition}%

@@

%%=============================================================%%
%% Prefix	-> \pfx{Dr}
%% GivenName	-> \fnm{Joergen W.}
%% Particle	-> \spfx{van der} -> surname prefix
%% FamilyName	-> \sur{Ploeg}
%% Suffix	-> \sfx{IV}
%% NatureName	-> \tanm{Poet Laureate} -> Title after name
%% Degrees	-> \dgr{MSc, PhD}
%% \author*[1,2]{\pfx{Dr} \fnm{Joergen W.} \spfx{van der} \sur{Ploeg} \sfx{IV} \tanm{Poet Laureate} 
%%                 \dgr{MSc, PhD}}\email{iauthor@gmail.com}
%%=============================================================%%

@@

%%==================================%%
%% sample for unstructured abstract %%
%%==================================%%

\abstract{
    Sensory areas of cortex respond more strongly to infrequent stimuli when these violate previously established regularities, a phenomenon known as deviance detection (DD). Previous modeling work has mainly attempted to explain DD on the basis of synaptic plasticity and tightly constrained network structure. However, a large fraction of cortical neurons also exhibit intrinsic plasticity, an underexplored potential mechanism, and recent results show DD in dissociated neuronal culture, raising doubts about the necessity of network structure.
    Here, we investigate DD in a spiking neuronal model with spatially constrained connectivity and two types of short-term plasticity, fast synaptic short-term depression (STD) and slower threshold adaptation (TA). We probe the model with oddball sequences at an intermediate inter-stimulus interval and assess DD.
    We find that TA alone is sufficient to generate DD signals and identify an interplay between suppression and its recovery as the underlying mechanism.
    Further, we find that the full model with both STD and TA yields a DD signal that exceeds the sum of the DD signals in ablated models with either STD or TA turned off. We identify the cause of this in an interaction where the fast-decaying STD increases local suppression near the frequently stimulated neurons, reducing the stimulus response, and thereby indirectly promotes global recovery of TA.
    We suggest that similar interactions between various forms of plasticity are likely to shape how sensory networks process regularities.
}

>>>>

@@

@@

@@

\abstract{
    Sensory areas of cortex respond more strongly to infrequent stimuli when these violate previously established regularities, a phenomenon known as deviance detection (DD). Previous modeling work has mainly attempted to explain DD on the basis of synaptic plasticity and tightly constrained network structure. However, a large fraction of cortical neurons also exhibit intrinsic plasticity, an underexplored potential mechanism.
    Here, we investigate DD in a spiking neuronal model with simple spatially constrained connectivity and two types of short-term plasticity, fast synaptic short-term depression (STD) and slower threshold adaptation (TA). We probe the model with oddball sequences at an intermediate inter-stimulus interval and assess DD.
    We find that TA alone is sufficient to generate DD signals and identify an interplay between suppression and its recovery as the underlying mechanism.
    Further, we find that the full model with both STD and TA yields a DD signal that exceeds the sum of the DD signals in ablated models with either STD or TA turned off. We identify the cause of this in an interaction between the contribution of STD to the local suppression of neurons sensitive to the frequently presented stimulus, and the recovery of TA-derived global suppression across the network.
    We suggest that similar interactions between various forms of plasticity are likely to shape how sensory networks process regularities.
}
